Foraging: The Triumphant Return

### 2022-05-05 manuscript.tex
Updates from Overleaf

Foraging, the act of searching for and gathering multiple targets (such as food), has been studied in both non-human animal contexts \cite{dawkins1971} and in human psychology studies \cite{kristjansson2014}. In many cases, the pattern of foraging behaviour has been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. However, we have recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022foraging}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2018, tagu2020}.

@@

We will use two previously published datasets to explore how well our model can account for trial-level behaviour. We will start with the visual foraging data from \cite{clarke2022qjep}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022qjep} used a desktop computer and mouse clicks). Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022foraging}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for a more details.

@@

We will make use of the foraging model from \cite{clarke2022foraging}. This treats foraging as a sampling without replacement process in which each item $i$ has probability $p_i$ of being selected as the next target. The $p_i$ depend on four parameters:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022qjep}. For each target selection, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the largest target manage to capture how well often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.png}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The *x*-axis gives the largest weight assigned by the model while the *y*-axis shows how often that target was actually selected by a human participant. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

>>>>

Foraging, the act of searching for and gathering multiple targets (such as food), has been studied in both non-human animal contexts \cite{dawkins1971} and in human psychology studies \cite{kristjansson2014}. In many cases, the pattern of foraging behaviour has been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. However, we have recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2022, tagu2020}.

@@

We will use two previously published datasets to explore how well our model can account for trial-level behaviour. We will start with the visual foraging data from \cite{clarke2022}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022} used a desktop computer and mouse clicks). Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for a more details.

@@

We will make use of the foraging model from \cite{clarke2022}. This treats foraging as a sampling without replacement process in which each item $i$ has probability $p_i$ of being selected as the next target. The $p_i$ depend on four parameters:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the largest target manage to capture how well often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.pdf}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The *x*-axis gives the largest weight assigned by the model while the *y*-axis shows how often that target was actually selected by a human participant. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_indiv_diff.pdf}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The *x*-axis gives the largest weight assigned by the model while the *y*-axis shows how often that target was actually selected by a human participant. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_ex_paths.pdf}
\caption{(\textit{top}) Two randomly selected trials in which the model does a good job in accounting for human behaviour. When participants diverge from the model's prediction, it appears to be due to some form of path-length optimisation. (\textit{bottom}) Two randomly selected examples in which the model does a less good job in accounting for the order in which human participants selected the items.}
\label{fig:qjep_paths}
\end{figure}
